Tree growth, mortality, and climatic stress in west-coast states, USA

### 2025-10-13 Manuscript.Rmd
Updates to MS.  Added current Intro and Discussion material, cleaned up figure legends for Figs 3 & 4

fig.diff_grow <- fig_nums(name = "fig_diffgrow", 
                          caption = paste0("Differences in mean growth rates (cm\u00B2/decade) between domains by species.  Domain abbreviations are described in the legend for ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are coniferous; those below are deciduous."))

fig.diff_mort <- fig_nums(name = "fig_diffmort", 
                          caption = paste0("Differences in mean decadal mortality rates between domains by species.  Domain abbreviations are described in the legend for ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are coniferous; those below are deciduous."))

@@

@@

We obtained empirical means and 95% confidence intervals via bootstrap. For every species we sampled plots with replacement and obtained point estimates.  For each domain we calculated a mean and upper and lower 95% CIs with the 2.5% and 97.5% quantiles from 1000 sampling iterations.  This approach allows for nonsymmetrical CIs around a mean and prevents CIs from including negative values.  We estimated contrasts between pairs of CWD domains by subtracting within-iteration domain point estimates.

Our analysis approach results in many comparisons.  We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates. The MC permutation analysis had 1000 iterations; for each iteration the bootstrap sampling was conducted 1000 times.  The domain assignments for each species was sampled without replacement.  The distance of contrast CI bounds from zero were compared against test CI bounds: The permutation recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds were.  The false detection rate for a

[permutation analysis]

@@

(Thoughts about permutation analysis): The number of significant findings per contrast was unexpected by chance except for Mortality, Above vs. Below.  We examined the likelihood of each of the significant comparisons arising by chance, adjusted by the number of significant findings (120).  Results were considered significant if their permutation p-value was below 0.00041 (0.05/120).  We might need to run 10,000 permutations.

Mortality: 
Picea engelmannii: Highest upper mortality = low. 
Abies lasiocarpa: Lowest mortaity in Low. 
- delta CWD lower for PIEN than for PILA.  Mean of High Above = 7.5 vs. 9ish.  
- Similar patterns of delta CWD vs. CWD
- ABLA = Least mortality in Low, in OLY penninsula, NW Cascades of WA.  High mortality overall (0.22)
- PIEN = most mort in NE WA. High mortality overall (0.28)
- PILA = most mort in mid/southern Sierras.  Seems like high mortality (0.20)

ABLA and PIEN are conspecifics / codominants. Perret et al. 2023 estimated rangewide mortality of ABLA of 25% +/-2.7, mort concentrated in eastern OR cascades (and ID and CO). 83% of mort due to fire or bio disturbance.  Conspecifics mostly affected by beetles, ABLA by fire.  PIEN = higher survival and longevity (used to be; they estimate around a high of 1% annual mortality for fir, below 0.5% annually in spruce, PIEN), ABLA = faster growing, more prolific regeneration (Andrus 2018).  We found that PIEN had higher mortality in the Above Low and Medium domains.  We note that the Above domains represents a proportionally large increase for Low and Medium, with CWD means of approximately 50 and 80 mm, respectively.  The same is true for ABLA, however, where the mean for Low is approximately 40 mm yet PIEN has less mortality in the Low domains and PIEN has more mortality in the Above Low domain.  The difference may be geographical: ABLA low-mortality plots are primarily on the Olympic Peninsula of Washington and the northwestern Cascades, where rainfall is abundant.  PIEN is generally absent from these locations.  However, PIEN has the least mortality in the Above domain in the Blue Mountains of Oregon and Washington while ABLA does not show a similar finding for the same geographical region (their AH domains largely overlap and exist only in that mountain range). PIEN mortality in N&NE WA may have been driven by spruce beetles (WADNR 2023)

Pinus contorta: Greatest mort in Low, least in High
Pinus ponderosa: Greatest mort in High
- In southern Sierra I see that they co-occur but have opposite mortality signals. Also a scattering in the Blue Mtns of opposite signals domains. Different fire strategies; Keely et al. 2011 may be useful.

Growth: 
PIEN: greater growth in High, above and below.  So, greater growth where least mortality.  
ABLA: not really any difference in growth by domain.  Supposed to be faster growing than PIEN, but not seeing it.

Lowest growth rate is for PIEN BL, at 15 in2.  Highest growth for ABLA is 12.5 in2. ABLA is supposed to be a faster grower (height) but maybe not DIA?  Diameter growth may be slow (trees 10-20" may be 150-200 yrs; Alexander et al. 1990).  On other hand, PIEN appears to have the capability of achieving greater diameters relatively sooner than ABLA (Parish et al. 1999).  PIEN growth in the High domains may be related to more growing days relative to other domains.

Communities:
Pinus lambertiana PILA
Pinus monticola PIMO
(Pinus ponderosa) PIPO
(Pinus contorta) PICO

These two are interesting because the lowest mortality domain for PIMO (Above, High) coincides geographically in the southern Sierra with the highest mortality domain for PILA (and for much of the highest-mortality domain for PIPO).  PILA has the highest growth in AH whereas there isn't much of a growth story for PIMA.  PIPO and PICO have great range overlap.  PICO has the highest mortality in Low (NE WA), particularly AL and lowest mortality in BH (Sierras, OR Cascades).  Note that the mortality rate for PICO in WA was the highest of any species in any state, and more than twice as high as the mortality rate for OR and 5 times higher than CA.  PICO growth is lowest in AL.  PIPO has the highest growth in Low.  PIPO in NE WA has much less acreage of pine beetle damage in 2023 than PICO, not really a factor for PIMO.

What is happening in the southern Sierras?
Southern Sierras: Paper by Restaino et al. 2019 examined mortality in Sierras.  2012-2016 CA drought most extreme in past 1000 yrs (citations) - hot drought (high temps, low precip).  Killed 129m trees in CA, primarily on west slope of Central & S Sierras (citation = USDA 2017).  In the areas of the greatest mortality, Fettig et al. (2019) found tree mortality rates of around 50% with PIPO exhibiting almost 90% mortality rates. Mortality was hightest for PIPO at the lowest elevations and attributable to western pine beetles, and concentrated in the largest diameter trees.  PILA was the second highest at 48%, and killed by Mt pine beetle. ABCO and CADE exhibited 26% and 23% mortality. Only one Quercus died.  Restaino et al. 2019 also found PIPO to experience high mortality followed by PILA, ABCO, and CADE. QUKE had 1% mortality (191 trees total, presumably 1 or 2 died?).  Large PIPO were most likely to die, esp in dense stands and dry areas. For PILA mortality positively affected by stand BA size. CADE survived much better in wetter sites.  ABCO mortality greater at dry sites, particularly for larger stands.  In our study we had no evidence of low mortality rates for QUKE - about 20%. We had some evidence of high mortality for ABCO in the Sierras, although not overwhelming.  CADE definitely had the highest mortality in the southern Sierras. PIMO appears to be doing great...why?

N/NE wA?  
McNellis et al. (2015) used FIA plot data to obtain mortality estimates but did not obtain high values for N/NE WA. Stanke (2021) shows a low forest stability index for PICO in that region. ABLA has a moderately negative index in that region. PIEN roughly aligns.  PIPO has agreeably low mortality for that region. PSME is high relative to coastal trees, although the differences are not great (maybe why Stanke thinks they're doing fine).   THPL is doing pretty well in NE WA; note that there is a call of alarm for the species (Andrus et al. 2023), even in that region.

Pseudotsuga menziesii PSME
So much data.  Over 4500 plots.  Except for High, where mortality of Above and Below was approximately equivalent, we see that mortality was high in AL and AM.  BL, BM had the lowest mortality.  These coincided with moister plots.  Growth was highest for Below except for BH.  PSME mortality appears to be drought related - see Cline et al. 2025. 
-Tree mortality increased 2020-2023, prolongued drought and warming.  Soil moisture, temperature, and VPD key predictors of old tree (125-500yr) mortality. 
-strongest signal at mid-elevation west of Cascade crest, OR
-Strongest for suppressed/intermediate, non-dominant or codominant individuals, smaller diameter (competition exacerbated the effect of climate variables on smaller-tree mortality at drier sites). 
-insect mortality more prevalent following drought. 
- lots of relevant citations
- Western hemlock (Tsuga heterophylla) in the S/I cohort also died more frequently.  ACMA also did at one site.  
- PNW heat dome of 2021 caused 11% mortality of C/D trees at Falls Creek site

Acer macrophyllum ACMA
We found that mortality was higher for BH than BL, BM. Maybe more interesting, growth was higher for Low, then Medium, then High irrespective of Above, Below. We cannot explain outright the high mortality for ACMA in California.  However, Betzen et al. (2021) found widespread evidence in Washington of ACMA decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and mortality associated with higher temperatures and water limitations. These findings are technically in alignment with the overall pattern in growth and mortality we observed, but does not specifically address the very high mortality rate of CA relative to OR and WA.

>>>>

fig.diff_grow <- fig_nums(name = "fig_diffgrow", 
                          caption = paste0("Differences in mean growth rates (cm\u00B2/decade) between domains by species.  Domain abbreviations are described in the legend for ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are coniferous; those below are deciduous. Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05, while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively."))

fig.diff_mort <- fig_nums(name = "fig_diffmort", 
                          caption = paste0("Differences in mean decadal mortality rates between domains by species.  Domain abbreviations are described in the legend for ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are coniferous; those below are deciduous.  Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05 while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively. "))

@@

There  are signs that tree species in western North America are in regional decline.  Johnston et al. (2025) document a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.  High-severity fires in California in 2020 and 2021 killed 13%-19% of giant sequoia (Sequoiadendron gianteum) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al. 2021); recruitment in these burns was depressed (Soderberg et al. 2022).   Five of eight common species in western North America exhibit declines in relative density  as measured by relative stand density (Stanke et al. 2021).  Shriver et al. (2022) found that four of five juniper and piñon pine species were declining in portions of their range, comparing mortality to recruitment rates.

Climate change as manifest through increases or intensification of drought and aridity can affect tree mortality directly through drought and indirectly by increasing wildfire frequency and severity and promoting insect and disease outbreak.  Hammond et al. (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.  Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt increasingly experienced in the western United States (Abatzoglou and Williams 2016, Westerling 2016, Juang et al. 2022).  There has also been a regional increase in the area of bark beetle outbreaks (Meddens et al. 2012, Hicke et al. 2016).  Many of these climate-influence mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.

Drought-induced mortality can be driven by carbon starvation and hydraulic failure (see reviews by Adams et al. 2010 and McDowell et al. 2008).  Carbon starvation occurs when isohydric species (e.g., the piñon pine group), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET) receive insufficient atmospheric carbon dioxide for photosynthesis.  Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in Populus tremuloides; Anderegg et al. 2015). Adams et al. (2017) performed a review of drought-induced mortality across 23 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.  Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al. 2022. Repetitive citation. Others?).

Drought and physiological stress from drought can indirectly increase mortality as well.  Drought stress can increase tree susceptibility to wildfire, insects, and disease.  Wildfire fire area (Abatzoglou and Williams, 2016; Parks et al. 2025) and number (Westerling 2016) are positively associated with increasing aridity in the western US.   Bark beetle outbreaks in semi-arid areas of the western US may be positively affected by or associated with acute intense drought and water-stressed trees (Negrón 2009, Gaylord et al. 2015) but chronically stressed trees may not provide like bark beetle populations with the resources sufficient to promote population expansion (Huberty and Denno, 2004).  Norway spruce (Picea abies) saplings inoculated with the necrotrophic pathogenic fungi Heterobasidion annosum sensu stricto and exposed to drought conditions were more likely to die than control saplings or saplings exposed to either drought conditions or inoculation, although Scots pine (Pinus sylvestris) seedlings did not show a similar relationship (Gomez-Gallego et al. 2022). Other insects and disease may be promoted or inhibited by drought, depending on the circumstances and due to conflicting factors (Kolb et al. 2016).

@@

We obtained empirical means and 95% confidence intervals via bootstrap. For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained point estimates for each species' domain.  For each domain we calculated a mean and upper and lower 95% CIs with the 2.5% and 97.5% quantiles from the  sampling iterations.  This approach allows for nonsymmetrical CIs around means.  We estimated contrasts between pairs of CWD domains by subtracting within-iteration domain point estimates.

Our analysis approach results in many comparisons.  We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values). The MC permutation analysis had 1000 iterations; each iteration relied on the bootstrap confidence interval estimation described above. The domain assignments for each species was sampled without replacement.  The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds. The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.

@@

The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S3 and S4), with values ranging from $p$ = 0.072 to $p$ < 0.0001 (41 instances).

# Discussion 
We detected numerous species for whom growth and/or mortality rates differed by CWD category.  We also found species for whom growth and mortality differed by state.  Our observed changes in CWD were not large, but regionally, from southern California to the Olympic Peninsula in Washington, CWD increased in 9 out of 10 plots.

Our regional probabilistically sampled design-based results provide a potentially valuable perspective on species status.  Acer macrophyllum growth was negatively associated with initial CWD, and its growth was lower and mortality almost three times higher in California than Washington and Oregon.  We cannot find primary literature that identifies an increase in mortality for this species in California; the degree of mortality observed seems severe.  However, Betzen et al. (2021) found widespread evidence in Washington of A. macrophyllum decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and increased mortality in areas associated with higher temperatures and water limitations.  Although these results align with our CWD findings for growth, they are not reflective (yet) of our state findings and are not apparent in our CWD findings for mortality.  We encourage a more intensive assessment of A. macrophyllum’s status and causes for decline in California.

A converse example is the mortality rate of Thuja plicata, where researchers and state agencies are concerned about die-off but our analysis did not find a strong mortality signal.  Andrus et al. (2023) sampled T. plicata at nine locations in Washington, northern Oregon, and Idaho that were without canopy dieback, exhibited canopy dieback, or had died. They found that the mortality was associated with exceptionally warm and dry climate conditions. A collaboration of state and federal agencies is currently reaching out to the public to identify dieback sites for this species (website: Goodrich et al. 2025).  Our analysis found that T. plicata exhibited the lowest decadal mortality rate and third highest growth rate of any species we examined and the greatest growth in Medium domains and some evidence that mortality in the Above Low domain (Olympic peninsula, western Washington Cascades) was greater than Above High domain (NE Washington, southern Oregon Cascades).  Our analysis may not be temporally sensitive enough to capture the same recent findings as Andrus et al., given that our FIA plot revisit data represented plot visits between 2011 to 2023.  However, our analysis does indicate that numerous species are exhibiting mortality rates at a state level or overall that are higher and are potentially unsustainable (e.g., Abies grandis, Abies lasiocarpa, Picea engelmannii, Pinus contorta, Pinus monticola, and Alnus rubra have state or overall decadal mortality rates greater than 0.25).

Abies lasiocarpa and Picea engelmannii are conspecifics and frequently codominant. Perret et al. (2023) estimated a rangewide mortality of A. lasiocarpa of 25% +/-2.7, with mortality concentrated in the eastern Oregon cascades as well as in Idaho and Colorado. 83% of the mortality was due to fire or biogenic disturbance.  These mortality rates are comparable to the state-level ones we found. Andrus et al. (2018) found that P. engelmannii exhibited higher survival and longevity in undisturbed plots in Colorado, with firs exhibiting mortalities of up to 1% annually while P. engelmannii had mortality rates below 0.5% annually.  We found that P. engelmannii had higher mortality in the Above Low domain.  We note that the Above domains represent a proportionally large increase for Low and Medium given initial CWD values. The same is true for A. lasiocarpa, yet P. engelmannii has greater mortality in the Low domains and A. lasiocarpa has less mortality in the Below Low domain although the MC permutation p-values were not very small.  The difference may be geographical: A. lasiocarpa low-mortality plots are primarily on the Olympic Peninsula of Washington and the northwestern Cascades, where rainfall is abundant.  P. engelmannii is generally absent from these locations.  However, P. engelmannii has the least mortality in the Above domain in the Blue Mountains of Oregon and Washington while A. lasiocarpa does not show a similar finding for the same geographical region. P. engelmannii mortality in northern and northeastern Washington, WA may have been driven by spruce beetles (WADNR 2023).
Our mortality estimate for Pinus contorta in Washington was twice that of Oregon’s and five times greater than Calfornia’s, while the growth confidence intervals for Washignton were lower than California’s and did not overlap. Mortality was highest in the Low CWD domains and growth lowest in the Above Low domain.  The Above Low domain aligns well with plots for northern and northeastern Washington.  Pine bark beetles (Dendroctonus ponderosae, Dendroctonus brevicomis, and Ips spp.) have affected this region for the past decade (WADNR 2023).  This area corresponds with the portions of P. contorta’s range in the lowest forest stability index category for Stanke et al. (2021) while California’s Sierra Nevadas and the Oregon and Washington Cascade Range correspond to positive stability index values.  Stanke et al. also found that insects (followed by fire) were a major driver of the percent change in relative tree density.

Our analysis centered around initial CWD and delta CWD as a variable of interest relative to tree growth and mortality.  There is ample evidence in the literature supporting the relationship or correlation of CWD or its cousin VPD to tree growth (Dyer 2004, Restaino et al. 2016) and mortality (Anderegg et al. 2015, Das et al. 2013, Hammond et al. 2022).  However, there are some challenges in relying on it in a design-based regional analysis.  We did not detect sizable changes in CWD between visits although CWD disproportionately increased between visits across sites.  The changes we observed may have represented greater proportional change for some species than others, with species such as Abies amablis, Alnus rubra, and Juniperus occidentalis exhibiting a low initial CWD range.

Interpreting CWD results at a regional scale becomes difficult because water stress alone can kill trees (Anderegg et al. 2015, Maherali et al.  2004, and van Nieuwstadt and Sheil 2005), but it can also increase the probability of insect (Hart et al. 2014, Preisler et al. 2012 , examples and counterexamples in Huberty and Denno 2004) and wildfire-caused mortality (Abatzoglou and Williams 2016). With reference to Manion’s (1991) “death spiral” (see also Bennett et al. 2023 for a CWD-focused version), teasing apart the cause of mortality from predisposing, inciting, and contributing factors is difficult at best, as these factors can interact (Gaylord et al. 2015).   Interrelated predisposing conditions for individual tree mortality such as canopy openness (Lemmo et al. 2024), structural complexity (Ma et al. 2023), tree height (Stovall et al. 2019), and within-stand competition (Ford et al. 2017) would be missed by our regional evaluation.
